---
output: html_document
editor_options: 
  chunk_output_type: console
execute: 
  cache: true
---

# calculating-marginal-effects

## Overview

::: callout-note
The following code is adapted from the [latitudinalCNDD repository](https://github.com/LisaHuelsmann/latitudinalCNDD/tree/main/code) by [Lisa Hüelsmann](https://demographicecology.com/).
:::

## Load libraries

```{r, message=FALSE}

# Load libraries
library(readr)
library(skimr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(parallel)
library(pbapply) # Adds progress bar to apply functions
library(here)
library(spatstat.geom)
library(mgcv) # For fitting gams
library(lubridate) # For calculating census intervals
library(broom) # For processing fit models
library(gratia)
library(boot)
library(mgcViz)
```

## Load data

```{r, message=FALSE}
data_BCI <- read_csv(here("./data/BCI_seedling data_30 species_2023_05_18.csv"))

```

# Visualize

```{r}
  
# Exploring data 

  # visualize 
   skim(data_BCI)
  ggplot(data_BCI, aes(x = con_dens)) +
  geom_histogram(binwidth = 1, color = "black", fill = "steelblue2") +
  labs(x = "Conspecific density", y = "Count", title = "Conspecific densities") + 
  theme_bw(12)
   
```

# Handling rare species

Species that have at least 4 unique Conspecific density values.

comment: In Lisa's code the species that did not fulfill these criteria and those where no convergence was achieved (overall 63.2 % of the species) were fitted jointly in one of two groups -- rare shrub species and rare tree species. she also has at least 20 alive and dead status observations each criteria.

```{r, message=FALSE, warning=FALSE}

# save census as factor
data_BCI$census <- factor(data_BCI$census)
data_BCI$spp <- factor(data_BCI$spp)
data_BCI$plot <- factor(data_BCI$plot)
data_BCI$s_height=data_BCI$height.last.census
data_BCI$interval=as.numeric(data_BCI$time.since.last.census)

# Handling rare species

nval = 4  ## this is the number of unique conspecific values you want

data_BCI %>% 
  group_by(spp) %>% 
  summarise(
            # ndead = sum(surv_next==0), # no needed now but may be useful to implement as criteria
            # nsurv = sum(surv_next==1),
            range_con_dens = max(con_dens) - min(con_dens),
            max_con_dens = max(con_dens),
            unique_con_dens = length(unique(con_dens)),
            unique_total_dens = length(unique(total_dens)),
            unique_height = length(unique(height.last.census))
  ) %>% 
  
  # add issues
  mutate(issue_nval = unique_con_dens < nval,                        # less than nval unique values in consp densities
         # issue_range = range_con_dens < minrange,                    # range should at least be equal to minrange
         # issue_cutoff = ndead < x | nsurv < x,                     # ndead or nsurv below cutoff
         trymodel = !(issue_nval),    # should a species specific model be tried at all? | issue_range | issue_cutoff
         rare = !trymodel                                          # preliminary assignment of rare species
  ) -> nsp

# we created a data frame  called "nsp"

```

## Function for fitting models

```{r, message=FALSE, warning=FALSE}
model_fit = function(data, speciesinfo, reduced = F) {
  
  # create new factor with correct factor levels per species (otherwise problem with margins)
  data$census = factor(data$census)
  
  
  # create model formula
  term_c = ifelse(length(unique(data$census)) > 1, " + s(census, bs = 're')", "") 
  
  if (reduced) {
    form =  as.formula(paste0("status ~ s(s_height, k = k1) + s(total_dens, k = k2)"
                              , term_c))
  } else {
    form =  as.formula(paste0("status ~ s(s_height, k = k1) + s(total_dens, k = k2)  + s(con_dens, k = k3)"
                              , term_c))
  }
  
   # chose penalty
  # set to default 10 (the same as -1)
  k1 = k2 = k3 = 10
  if (k1 > speciesinfo$unique_height) k1 = speciesinfo$unique_height - 2
  if (k2 > speciesinfo$unique_total_dens) k2 = speciesinfo$unique_total_dens - 2
  if (k3 > speciesinfo$unique_con_dens) k3 = speciesinfo$unique_con_dens - 2
  
  
  # fit model
  # https://stats.stackexchange.com/questions/27586/spline-df-selection-in-a-general-additive-poisson-model-problem/71300#71300
  mod = try(gam(form
                , family = binomial(link=cloglog)
                , offset = log(interval)
                , data = data
                # , method = "GCV.Cp" # tends to undersmooth
                , method = "REML"
  ) , silent = T
  )
  
  return(mod)
  
}

model_convergence = function(model) {
  
  # gam not available
  if (!any(class(model)=="gam")) {
    print(paste(spp, "gam failed"))
  } else {
    
    # gam not converged
    if (!model$converged) {
      print(paste(spp, "no convergence"))
    } else {
      
      # check for complete separation
      # https://stats.stackexchange.com/questions/336424/issue-with-complete-separation-in-logistic-regression-in-r
      # Explore warning "glm.fit: fitted probabilities numerically 0 or 1 occurred"
      eps <- 10 * .Machine$double.eps
      glm0.resids <- augment(x = model) %>%
        mutate(p = 1 / (1 + exp(-.fitted)),
               warning = p > 1-eps,
               influence = order(.hat, decreasing = T))
      infl_limit = round(nrow(glm0.resids)/10, 0)
      # check if none of the warnings is among the 10% most influential observations, than it is okay..
      num = any(glm0.resids$warning & glm0.resids$influence < infl_limit)
      
      # complete separation
      if (num) {
        print(paste(spp, "complete separation is likely"))
      } else {
        
        # missing Vc
        if (is.null(model$Vc)) {
          print(paste(spp, "Vc not available"))
        } else {
        
        # successful model
        return(model)
        }
      }
    }
  }
}

```

# Fit models for All (expect rare species)

```{r, message=FALSE, warning=FALSE}
res_mod = list()      # main model fits
res_red_mod = list()  # reduced model fits for Pseudo R2


res_mod = list()      # main model fits
res_red_mod = list()  # reduced model fits for Pseudo R2


# Fit models for individual species
for (spp in nsp$spp[nsp$trymodel]) {
  
  # select data for individual species
  dat_sp = data_BCI[data_BCI$spp == spp, ]
  
  # model fit and reduced fit for Pseudo R2
  mod = model_fit(data = dat_sp, speciesinfo = nsp[nsp$spp == spp, ])
  mod_red = model_fit(data = dat_sp, speciesinfo = nsp[nsp$spp == spp, ], reduced = T)
  
  # check model success
  res = model_convergence(model = mod)
  res_red = model_convergence(model = mod_red)
  
  # save result
  if (is.character(res)) {
    nsp$rare[nsp$spp == spp] = T  
  } else {
    res_mod[[spp]] = res
    res_red_mod[[spp]] = res_red
  }
}

```

# Fit models for rare species - Still a work in progress (Skip)

```{r, message=FALSE, warning=FALSE, eval=FALSE}

# overview rare versus rare and no convergence
table(rare = nsp$rare, trymodel = nsp$trymodel) # N: We have 9 species here that are considered rare

# make Rare dependent on stature - N: as Shrub or trees 
stature <- data.frame(spp = unique(data_BCI$spp), stature = "seedling") # we added this for seedlings but is about size
## for seedlings we should just name all of then species rare_seelings and add to the analyses.

nsp$stature = stature$stature[match(nsp$spp, stature$sp)]
nsp$rare_stature = NA
nsp$rare_stature[nsp$rare] = paste0("Rare_", nsp$stature[nsp$rare])


# explore available data per rare species group
data_BCI %>% 
  mutate(spp = nsp$rare_stature[match(spp, nsp$spp)]) %>% 
  filter(!is.na(spp)) %>% 
  group_by(spp) %>% 
  summarise(
            # ndead = sum(surv_next==0),
            # nsurv = sum(surv_next==1),
            range_con_dens = max(con_dens) - min(con_dens),
            max_con_dens = max(con_dens),
            unique_con_dens = length(unique(con_dens)),
            unique_total_dens = length(unique(total_dens)),
            unique_height = length(unique(height.last.census))
  ) %>% 
  
  # add issues
  mutate(issue_nval = unique_con_dens < nval,                            # less than nval unique values in conspecific densities
         # issue_range = range_con_dens < minrange,                         # range should at least be equal to minrange
         # issue_cutoff = ndead < x | nsurv < x,                          # ndead or nsurv below cutoff
         trymodel = !(issue_nval ),         # should a group specific model be tried at all?
  ) -> nsp_rare



---

# Fit models for species groups
for (spp in unique(nsp_rare$spp[nsp_rare$trymodel])) {
  
  # select data for all species in group
  sps = nsp$spp[which(nsp$rare_stature == spp)]
  dat_spp = data_BCI[data_BCI$spp %in% sps, ]
  
  # model fit
  mod = model_fit(data = dat_spp, speciesinfo = nsp_rare[nsp_rare$spp == spp, ])
  mod_red = model_fit(data = dat_spp, speciesinfo = nsp_rare[nsp_rare$spp == spp, ], reduced = T)  
  
  # check model success
  res = model_convergence(model = mod)
  res_red = model_convergence(model = mod_red)
  
  # save result
  if (!is.character(res)) {
    res_mod[[sp]] = res
    res_red_mod[[sp]] = res_red
  }
}


```

# Regression table via broom::tidy() --------------------------------------

```{r}
coefs = lapply(res_mod, broom::tidy)
coefs = Map(cbind, coefs, sp = names(coefs))
coefs = do.call(rbind, coefs)

```

# Model summary via broom::glance()

```{r}
# df logLik AIC BIC deviance df.residuals nobs 
sums = lapply(res_mod, broom::glance)
sums = Map(cbind, sums, sp = names(sums))
sums = do.call(rbind, sums)


# AUC
aucs = lapply(res_mod, function(x) {
  roc <- performance::performance_roc(x, new_data = x$model)
  bayestestR::area_under_curve(roc$Spec, roc$Sens)
})
sums$AUC = unlist(aucs)


# Pseudo R2
sums$pseudoR2 = 1 - (unlist(lapply(res_mod, function(x) x$deviance)) / unlist(lapply(res_red_mod, function(x) x$deviance)))
```

## AMEs absolute and Relative

AMEs relative are used to compare in meta-analyses while absolute AMEs are reported as they are commonly used in other studies.

We have will get here a file named res_model that have all the results and coefficients

## Function to calculte AMEs without the "gratia" package

```{r}
library(MASS)


# Useful functions --------------------------------------------------------


# set step on x for numerical derivative
setstep = function(x) {
  eps = .Machine$double.eps
  x + (max(abs(x), 1, na.rm = TRUE) * sqrt(eps)) - x
}



# Average marginal effects ------------------------------------------------

get_AME = function(mod, data, term
                   , change = NULL
                   , at = NULL
                   , offset = 1
                   , relative = F
                   , iterations = 1000
                   , seed = 10
                   , samples = F) {
  
  # Two data frames with shift in term of interest:
  d0 = d1 = data
  
  # if change is NULL, use numerical derivative
  if (is.null(change)) {
    
    d0[[term]] = d0[[term]] - setstep(d0[[term]])
    d1[[term]] = d1[[term]] + setstep(d1[[term]])
    
  } 
  
  # if change includes "+", use explicit additive change
  if (grepl("\\+", paste(change, collapse = "_"))) {
    
    d1[[term]] = d1[[term]] + as.numeric(gsub("\\+", "", change))
    
  } 
  
  # if change has two values, use explicit change
  if (length(change) == 2) {
    
    d0[[term]] = as.numeric(change[1])
    d1[[term]] = as.numeric(change[2])
    
  }
  
  # Fix at values:
  if (!is.null(at)) {
    for (i in names(at))
      d0[[i]] = at[[i]]
      d1[[i]] = at[[i]]
  }
  
  # Matrices for prediction, map coefs to fitted curves
  Xp0 <- predict(mod, newdata = d0, type="lpmatrix")
  Xp1 <- predict(mod, newdata = d1, type="lpmatrix")
  
  # Model settings
  ilink <- family(mod)$linkinv
  beta <- coef(mod)
  # vc <- mod$Vp # Bayesian
  vc <- mod$Vc # Bayesian accounting for smoothing parameter uncertainty
  

  # marginal effects
  pred0   <- 1 - (1-ilink(Xp0 %*% beta))^offset
  pred1   <- 1 - (1-ilink(Xp1 %*% beta))^offset
  ME <- (pred1-pred0)
  
  # if change is NULL, use numerical derivative
  if (is.null(change)) {
    ME <- ME/(d1[[term]] - d0[[term]])
  } 

  
  # convert to relative if requested
  if (relative == T) ME = ME/pred0
  
  # # Alternative for relative:
  # ME1 = ME/pred0; ME2 = ME/mean(pred0)
  # plot(ME, ME2)
  # points(ME, ME1, col = 2)
  # par(mfrow = c(1, 2))
  # hist(ME1); hist(ME2)
  # par(mfrow = c(1, 1))
  
  # average marginal effect
  AME = mean(ME)
  
  
  # generate AME samples
  
  # variance of average marginal effect via "posterior" simulation
  # simulate from multivariate normal using model beta means and covariance matrix
  if (!is.null(seed)) set.seed(seed)
  coefmat = mvrnorm(n = iterations
                    , mu = beta
                    , Sigma = vc)
  
  # estimate AME from each simulated coefficient vector
  AMEs = apply(coefmat, 1, function(coefrow) {
    
    # marginal effects
    pred0   <- 1 - (1-ilink(Xp0 %*% coefrow))^offset
    pred1   <- 1 - (1-ilink(Xp1 %*% coefrow))^offset
    ME <- (pred1-pred0)
    
    # if change is NULL, use numerical derivative
    if (is.null(change)) {
      ME <- ME/(d1[[term]] - d0[[term]])
    } 
    
    # convert to relative if requested
    if (relative == T) ME = ME/pred0
    
    # average marginal effect
    AME = mean(ME)
    return(AME)
  })
  
  # Combine results
  if (!samples) {
    res = data.frame(term
                     , estimate = AME
                     , std.error = sqrt(var(AMEs))  # not a good idea
                     , estimate.sim = mean(AMEs)    # not a good idea
                     , offset
                     , change.value = paste(change, collapse = "_"))
    return(res) 
    
  } else {
    
    res_sums = data.frame(term
                     , estimate = AME
                     , std.error = sqrt(var(AMEs)) # not a good idea
                     , offset
                     , change.value = paste(change, collapse = "_"))
    
    res_samples = data.frame(term
                             , estimate = AMEs
                             , MLE = AME
                             , offset
                             , change.value = paste(change, collapse = "_"))
    res = list(res_sums, res_samples)
    return(res)  
    
  }
}



```

## Settings for AMEs

Need to solve the value of additive is it 1?

```{r}

#### chose predictors for local density
  
predictors <- c(con_dens = "con_dens", 
                total_dens = "total_dens")

# change in conspecific density for AME calculations----
additive=1 # is 1 ? in the tree analyses was pi*((dbh_neighbor/1000)/2)^2 *  
#additive = pi*((dbh_neighbor/1000)/2)^2 *         # basal area (m2) of one more neighbor with dbh_neighbor
  # dec_fun(decay_con, dist_neighbor, decay_type)   # at a distance of dist_neighbor  

  
# fixed value for interval (offset)
# different change settings for conspecific densities
# averaged over all other predictors (total density, dbh...)

interval = 1
change = list(equilibrium = data.frame(con_dens = "paste('+', additive)")
              , invasion = data.frame(con_dens = "c(0, additive)")
              , iqr = data.frame(con_dens = "c(q1, q3)")
              )
iter = 500

```

# calculating AMEs

## Absolute AMEs values

```{r}
# Absolute AMEs -----------------------------------------------------------

# Calculate absolute AMEs based on manual function get_AME
AME = data.frame()
AMEsamples = data.frame()
for (i in names(predictors)[grepl("con_", names(predictors))]) { # for all predictors
  for (j in names(change)) {
    temp = lapply(res_mod, function(x){
      if (j == "iqr") {
        q1 = quantile(x$model$con_dens, probs = 0.25)
        q3 = quantile(x$model$con_dens, probs = 0.75)
      }
      get_AME(x
              , data = x$model
              , offset = interval
              , term = i
              , change = eval(parse(text = change[[j]][,i]))
              , iterations = iter
              , samples = T
      )
    }
    )
    
    # AME
    tempAME = lapply(temp, function(x) x[[1]])
    tempAME = Map(cbind, tempAME, change = j, sp = names(tempAME))
    tempAME = do.call(rbind, tempAME)
    AME = rbind(AME, tempAME)
    
    # AME samples
    tempSamples = lapply(temp, function(x) x[[2]])
    tempSamples = Map(cbind, tempSamples, change = j, sp = names(tempSamples), iter = iter)
    tempSamples = do.call(rbind, tempSamples)
    AMEsamples = rbind(AMEsamples, tempSamples)
  }
}
```

## Relative AMEs values

```{r}
# Relative AMEs -----------------------------------------------------------


# Calculate relative AMEs based on manual function get_AME
rAME = data.frame()
rAMEsamples = data.frame()
for (i in names(predictors)[grepl("con_", names(predictors))]) { # for all predictors
  for (j in names(change)) {
    temp = lapply(res_mod, function(x){
      if (j == "iqr") {
        q1 = quantile(x$model$con_dens, probs = 0.25)
        q3 = quantile(x$model$con_dens, probs = 0.75)
      }
      get_AME(x
              , data = x$model
              , offset = interval
              , term = i
              , change = eval(parse(text = change[[j]][, i]))
              , iterations = iter
              , relative = T
              , samples = T
      )
    }
    )
    
    # rAME
    tempAME = lapply(temp, function(x) x[[1]])
    tempAME = Map(cbind, tempAME, change = j, sp = names(tempAME))
    tempAME = do.call(rbind, tempAME)
    rAME = rbind(rAME, tempAME)
    
    # rAME samples
    tempSamples = lapply(temp, function(x) x[[2]])
    tempSamples = Map(cbind, tempSamples, change = j, sp = names(tempSamples), iter = iter)
    tempSamples = do.call(rbind, tempSamples)
    rAMEsamples = rbind(rAMEsamples, tempSamples)
  }
}

```

# Plotting and saving results

```{r}
# plot splines in pdf -----------------------------------------------------

# Set the PDF file name
pdf_file <- here("./plots/mortality.pdf")

# Set the PDF device to save the plot
pdf(pdf_file)


for (i in 1:length(res_mod))  {
  
  # with mgcv
  # plot(res_mod[[i]], pages = 1, scale = 0, main = names(res_mod)[i], all.terms = T) 
  
  # with gamViz
  vizmod <- getViz(res_mod[[i]], post = T, unconditional = T)
  pl = plot(vizmod, nsim = 20, allTerms = T) + 
    l_ciLine() + l_fitLine() + l_simLine() + 
    l_ciBar() + l_fitPoints(size = 1) + 
    l_rug() + 
    # l_points() +
    labs(title = names(res_mod)[i]) 
  print(pl, pages = 1)
  
}
dev.off()


# Save results ------------------------------------------------------------


save(list = c("AME", "AMEsamples", "rAME", "rAMEsamples", "nsp", "coefs", "sums") # "nsp_rare"
     , file = paste0(here("./data/mortality.Rdata")))
load(here("./data/mortality.Rdata"))

```

## Fit one species (Skip)

```{r, eval = FALSE}

#Fit only one specie
## interval should be the time betwen censuses in years###
data_BCI_1=data_BCI[data_BCI$spp=="QUARAS",]
mod <- gam(formula = status ~ s(s_height, k = k1) + s(total_dens, k = k2) + s(con_dens, k = k3),
               family = binomial(link = cloglog),
               offset = log(interval),
               data = data_BCI_1,
               method = "REML")

y_d <- response_derivatives(mod, data = data_BCI_1, type = "central",
                            focal = "con_dens", eps = 0.01, seed = 21)
derivative_values <- y_d$derivative
mean_derivative <- mean(derivative_values)
#also calculate CI

# draw response derivatives
p2 <- y_d |>
    ggplot(aes(x = con_dens, y = derivative)) +
    geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.2) +
    geom_line() +
    labs(title = "Estimated 1st derivative of estimated count",
         y = "First derivative")

```
